# Peer-graded Assignment: Course Project 1

```{r Run Date, include=FALSE}
Runtime <- Sys.Date()
```

---
title: "PA1_template.Rmd"
author: "Julian Buhagiar"
date: 2017-04-07
output: html_document
---

date last modified: `r Runtime` 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# remove runtime used which is only used for the header of this document
rm(Runtime)
```



## Questions

      •[What is mean total number of steps taken per day?][1]
      
      •[What is the average daily activity pattern?][2]
      
      •[Are there differences in activity patterns between weekdays and weekends?][3]


## Description

This assignment makes use of data from a personal activity monitoring device. 
This device collects data at 5 minute intervals through out the day. The data 
consists of two months of data from an anonymous individual collected during 
the months of October and November, 2012 and include the number of steps taken 
in 5 minute intervals each day.
```{r source file, echo=TRUE}
URL = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
```

## Variables

## System Info and Library Prerequisites

### System

### Libraries

### Prerequisite Common Functions

#### Workspace Housekeeping

#### Retrieve data files

## Analysis Run Sequence

### Question 1 - What is mean total number of steps taken per day?

### Question 2 - What is the average daily activity pattern?

### Question 3 - Are there differences in activity patterns between weekdays and weekends?

[1]: https://github.com/JulesBuh/RepData_PeerAssessment/blob/master/PA1_template.Rmd#question-1---what-is-mean-total-number-of-steps-taken-per-day

[2]: https://github.com/JulesBuh/RepData_PeerAssessment/blob/master/PA1_template.Rmd#question-2---what-is-the-average-daily-activity-pattern

[3]: https://github.com/JulesBuh/RepData_PeerAssessment/blob/master/PA1_template.Rmd#question-3---are-there-differences-in-activity-patterns-between-weekdays-and-weekends