---
output: 
  html_document: 
    keep_md: yes
    toc: yes
---
# Peer-graded Assignment: Course Project 1

```{r Run Date, include=FALSE}
Runtime <- Sys.Date()
```

---
title: "PA1_template.Rmd"
author: "Julian Buhagiar"
date: 2017-04-07
output: html_document
---

date last modified: `r Runtime` 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# remove runtime used which is only used for the header of this document
rm(Runtime)
```



## Questions

[What is mean total number of steps taken per day?][1]
      
[What is the average daily activity pattern?][2]
      
[Are there differences in activity patterns between weekdays and weekends?][3]


## Description

This assignment makes use of data from a personal activity monitoring device. 
This device collects data at 5 minute intervals through out the day. The data 
consists of two months of data from an anonymous individual collected during 
the months of October and November, 2012 and include the number of steps taken 
in 5 minute intervals each day.
```{r source file, echo=TRUE}
URL = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
```

## Variables

included in the source dataset are:

• `steps`: Number of steps taking in a 5-minute interval (missing values are coded as NA)

• `date`: The date on which the measurement was taken in YYYY-MM-DD format

• `interval`: Identifier for the 5-minute interval in which measurement was taken

## System Info and Library Prerequisites

### System
```{r prerequisites, echo=FALSE}

      message("R version\t",version$version.string)
      message("OS\t",version$os)
```
### Libraries
```{r load libraries, include=FALSE}
# 1.0 prerequisite
      library(dplyr)
      library(stringr)
```

```{r library status, echo=FALSE}
# 1.0 prerequisite
      if(!exists("tbl_df")){
            message<-message("dplyr 0.4.0 or greater is required")
            return()
      }
      
      message("dplyr ",packageVersion("dplyr")," was used during for the production of this analysis")
      if(!exists("str_pad")){
            message<-message("stringr 1.1.0 or greater is required")
            return()
      }
      
      message("stringr ",packageVersion("stringr")," was used during for the production of this analysis")

```
### Prerequisite Common Functions

#### Workspace Housekeeping

`keep()`is an internal function which keeps a record of the important variables 
that the script will retain in the workspace following the running of the code.

`cleanWorkspace()` is a finalising function which removes the `<<-` variables that 
aren't intended to be retained after the running of the function.
```{r keep, include=TRUE}
keep<-function(...){
      #>DESCRIPTION----
      # keeps a record of starting globals
      # and appends the desired variables that we wish to keep by the end of the 
      # script, thereby making all other variables and functions that are made during 
      # the running of this function as temporary.
      if(!exists("GlobalEnvKeep")){
            GlobalEnvKeep<<-c(ls(envir = as.environment(globalenv())))
      }
      GlobalEnvKeep<<-c(GlobalEnvKeep,...)
}
keep() 
cleanWorkspace<-function(){
      rm(list= ls(envir=as.environment(globalenv()))[!(ls(envir=as.environment(globalenv())) %in% GlobalEnvKeep )],
         envir=as.environment(globalenv()))
      rm(keep,envir=as.environment(globalenv()))
}
```

#### Retrieve data files

`sourceData()` is used to download the data to the workspace and the working folder.

```{r sourceData function, include=TRUE}
sourceData<-function(src=URL,expectedfiles="activity.csv",workspaceNames="sourcedData"){
      
      
      if(!file.exists(expectedfiles)){
            message("downloading files")
            download.file(src,"data.zip")
            unzip("data.zip")
            file.remove("data.zip")
      }
      
      columnClassPreDefined<-c("numeric","character","character")
      ## This first line will likely take a few seconds to load
      if (!exists(workspaceNames)){sourcedData<<- read.csv(expectedfiles,
                                                           colClasses=columnClassPreDefined,
                                                           header = TRUE,
                                                           stringsAsFactors = FALSE,
                                                           na.strings = "NA")
      #change the date column to a date type field
      sourcedData$date<<-as.Date(sourcedData$date)
      #pads the interval to become a 4 digit time
      sourcedData$interval<<-stringr::str_pad(sourcedData$interval,width=4,side="left",pad="0")
      # combines data and time into a single column called time
      sourcedData$time<<-paste(sourcedData$date,sourcedData$interval)
      sourcedData$time<<-as.POSIXct(strptime(sourcedData$time,"%Y-%m-%d %H%M"))
      sourcedData$timeOnly<<-as.POSIXct(strptime(sourcedData$interval,"%H%M"))
      ## retains the sourcedata variable for the project workspace
      keep("sourcedData")
      }
      # organising data----
      sourcedData<<-dplyr::tbl_df(sourcedData)
      
      message("Data loaded successfully into the workspace as 'sourcedData'")
      
      ##creates a container to keep useful metadata as a list
      metadata<<-list(NULL)
      metadata$head<<-(head(sourcedData))
      metadata$summary<<-summary(sourcedData)
      str(sourcedData)
      
}

```

## Analysis Run Sequence

The call to the function `sourceData()` loads the data into R.

```{r load data}

sourceData(URL)
```

The data when it was loaded in shows there are `r nrow(sourcedData)` observations 
and `r length(sourcedData)-1` variables. The `interval` variable represents a 24 hour time
and a new variable has been introduced called `time` joining the date and time together into a 
single POSIXct date time. The variables are called `r names(sourcedData)`.

Running `summary()` on the loaded data shows there are `r metadata$summary[7,1]`
The observations were taken between `r range(sourcedData$date)`.

### Question 1 - What is mean total number of steps taken per day?

### Question 2 - What is the average daily activity pattern?

### Question 3 - Are there differences in activity patterns between weekdays and weekends?

[1]: https://github.com/JulesBuh/RepData_PeerAssessment/blob/master/PA1_template.Rmd#question-1---what-is-mean-total-number-of-steps-taken-per-day

[2]: https://github.com/JulesBuh/RepData_PeerAssessment/blob/master/PA1_template.Rmd#question-2---what-is-the-average-daily-activity-pattern

[3]: https://github.com/JulesBuh/RepData_PeerAssessment/blob/master/PA1_template.Rmd#question-3---are-there-differences-in-activity-patterns-between-weekdays-and-weekends